CITK - an architecture and examples of CUDA enabled ITK filters

### 2011-05-17 article/Article.tex
minor article updates

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  hyperref should be the last package to be loaded.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[dvips,
bookmarks,
bookmarksopen,
backref,
colorlinks,linkcolor={blue},citecolor={blue},urlcolor={blue},
]{hyperref}
% to be able to use options in graphics
\usepackage{graphicx}
% for pseudo code
\usepackage{listings}
% subfigures
\usepackage{subfigure}

@@

% At minimum, give your name and an email address.  You can include a
% snail-mail address if you like.
\author{Richard Beare{$^1$}, Daniel Micevski, Chris Share\\Luke Parkinson, Phillip Ward, Mike Kuiper{$^2$}}
\authoraddress{{$^1$}Richard.Beare@monash.edu, Monash University, Melbourne, Australia\\
{$^2$}mike@vpac.org, Victorian Partnership for Advanced Computing, Melbourne, Australia.}

@@

\section{CITK Architecture}
The aim of the architecture outline below was to allow GPU enabled
filters to be included in an application without change of programming
style or losing performance via redundant host to device memory copies.

@@

The approach used in CITK does require a modification to core ITK
classes, but has a number of advantages. A similar approach has since
been outlined on the ITK Wiki {\url http://www.cmake.org/Wiki/ITK_Release_4/GPU_Acceleration}.

@@

\subsection{Weaknesses}
\begin{itemize}
\item This framework only supports CUDA, and not OpenCL. The ITK 4.0
proposal supports OpenCL. Limitations of the CUDA development
environment mean that even CUDA integration is less complete than
hoped, with significant changes to compilation processes being
necessary (see below).
\item A copy between host and device always results in the source of
the copy being considered redundant. This could be inefficient in some
cases. The problem is correctly dealing with identical copies on both
host and device. If, for example, a pipeline is branched such that one
branch in on GPU and the other on CPU, then the branch point is likely
to become a source of redundant copies.
\item The need to copy between device and host memory breaks some of
the usual assumptions, leading to some ugly use of {\em mutable}
declarations in CudaImportImageContainer.
\item Morphology filters included in the package use texture memory
but are not as generic as the standard ITK versions.
\item ITK filters are able to provide their own implementations of key
methods, such as {\em AllocateOutputs}. This can lead to problems when
connecting CUDA enabled components to a CPU pipeline. This problem has
been observed in the {\em itkStatisticsImageFilter}, which is a
multi-threaded {\em ImageToImageFilter} with its own AllocateOutputs
method passing input through to output. This bypasses the trigger to
copy device memory back to the host, leading to incorrect results.
\end{itemize}

@@

\begin{itemize}
\item {\url http://voxel.dl.sourceforge.net/sourceforge/itk/InsightToolkit-3.20.0.tar.gz}
\item extract
\item cd ITK-3.20
\item patch -p0 < path/to/patch.3.20.0.dif
\end{itemize}

Alternatviely, this code may currently be retrieved via git, as follows:
\begin{itemize}
\item git clone git://github.com/richardbeare/ITK.git
\item cd ITK
\item git checkout v3.20.0
\item git branch v3.20.0_cuda
\end{itemize}

@@

\subsection{Build examples}
\subsection{Workarounds for nvcc weaknesses/bugs}
\begin{itemize}
  \item multiple include of SSE files - turn off SSE options for VNL - advanced/VNL.
\end{itemize}

@@

@@

The corresponding CudaAddImageFilterKernel.cu file contains:
\begin{verbatim}
template <class T, class S>
__global__ void AddImageKernel(T *output, const S *input, int N)
{
  int idx = blockIdx.x * blockDim.x + threadIdx.x;
  if (idx<N) 
    {
    output[idx] += input[idx];
    }
}

@@

\end{verbatim}

This is a simple structure that allows a number kernels to be precompiled for different voxel types.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  hyperref should be the last package to be loaded.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[dvips,
bookmarks,
bookmarksopen,
backref,
colorlinks,linkcolor={blue},citecolor={blue},urlcolor={blue},
]{hyperref}
% to be able to use options in graphics
\usepackage{graphicx}
% for pseudo code
\usepackage{listings}
% subfigures
%\usepackage{subfigure}

@@

% At minimum, give your name and an email address.  You can include a
% snail-mail address if you like.
\author{Richard Beare{$^1$}, Daniel Micevski, Chris Share\\Luke
  Parkinson, Phillip Ward, Wojtek Goscinski{$1$}, Mike Kuiper{$^2$}}
\authoraddress{{$^1$}Richard.Beare@monash.edu, Monash University, Melbourne, Australia\\
{$^2$}mike@vpac.org, Victorian Partnership for Advanced Computing, Melbourne, Australia.}

@@

\section{CITK Architecture}
The aim of the architecture outlined below was to allow GPU enabled
filters to be included in an application without change of programming
style or losing performance via redundant host to device memory copies.

@@

The approach used in CITK does require a modification to core ITK
classes, but has a number of advantages. A similar approach has since
been outlined on the ITK Wiki \url{http://www.cmake.org/Wiki/ITK\_Release\_4/GPU\_Acceleration}.

@@

Some exceptions have been uncovered during this development. Minor
changes have also been made to the {\em AllocateOutputs} method of the
{\em itkInPlaceImageFilter} to support pipelines that connect in
place, multi-threaded CPU filters to a CUDA filter. Other exceptions
are discussed below.

\subsection{Weaknesses}
\begin{itemize}
\item This framework only supports CUDA, and not OpenCL. The ITK 4.0
proposal supports OpenCL. Limitations of the CUDA development
environment mean that even CUDA integration is less complete than
hoped, with significant changes to compilation processes being
necessary (see below).
\item A copy between host and device always results in the source of
the copy being considered redundant. This could be inefficient in some
cases. The problem is correctly dealing with identical copies on both
host and device. If, for example, a pipeline is branched such that one
branch in on GPU and the other on CPU, then the branch point is likely
to become a source of redundant copies.
\item The need to copy between device and host memory breaks some of
the usual assumptions, leading to some ugly use of {\em mutable}
declarations in CudaImportImageContainer.
\item Morphology filters included in the package use texture memory
but are not as generic as the standard ITK versions.
\item ITK filters are able to provide their own implementations of key
methods, such as {\em AllocateOutputs}. This can lead to problems when
connecting CUDA enabled components to a CPU pipeline. This problem has
been observed in the {\em itkStatisticsImageFilter}, which is a
multi-threaded {\em ImageToImageFilter} with its own AllocateOutputs
method passing input through to output. This bypasses the trigger to
copy device memory back to the host, leading to incorrect
results. Other filters with unusual structure are likely to cause
problems. The simple fix for such filters is to call {\em
  GetBufferPointer} in the {\em AllocateOutputs} method.
\end{itemize}

@@

\begin{itemize}
\item \url{http://voxel.dl.sourceforge.net/sourceforge/itk/InsightToolkit-3.20.0.tar.gz}
\item extract
\item cd ITK-3.20
\item patch -p0 $<$ path/to/patch.3.20.0.dif 
\end{itemize}

Alternatviely, this code may currently be retrieved via git, as follows:
\begin{itemize}
\item git clone git://github.com/richardbeare/ITK.git
\item cd ITK
\item git checkout v3.20.0\_cuda
\end{itemize}

@@

\subsection{Build examples}
The examples and tests are included in this package. The most up to
date version of the package can be found at \url{http://code.google.com/p/cuda-insight-toolkit/}.

@@

Pointers to device memory are obtained using the {\em
  GetDevicePointer} method and are passed to CUDA kernel functions.

@@

The corresponding CudaAddImageFilterKernel.cu file contains:
{\small
\begin{verbatim}
template <class T, class S>
__global__ void AddImageKernel(T *output, const S *input, int N)
{
  int idx = blockIdx.x * blockDim.x + threadIdx.x;
  if (idx<N) 
    {
    output[idx] += input[idx];
    }
}

@@

\end{verbatim}
}
This is a simple structure that allows a number kernels to be precompiled for different voxel types.

\subsection{Using {\em thrust} algorithms}
The thrust project, \url{http://code.google.com/p/thrust/}, is a
source of templated, CUDA-enabled algorithms. The
CudaStatisticsImageFilter makes use of these algorithms. It is also
possible to implement simple arithmetic filters using the thrust {\em
  transform} algorithm (leading to more elegant code), but preliminary
tests suggest a significant loss in performance. There are
examples of thrust-based arithmetic in several sample filters that can
be switched on with a cmake option.

\section{Testing}
The online testing within the Insight Journal does not support CUDA
and therefore cannot be used to test this contribution.
\subsection{ITK}
Changes to ITK classes have been tested using standard ITK tests,
producing the same results as an unmodified ITK.
\subsection{Testing CUDA filters}
A range of simple CUDA enabled filters have been developed and
compared to the CPU equivalents. CMake based test are included in this
contribution.

\section{Performance}
Improved computational performance is the reason for interest in GPU
imaging applications. However there are many stories of how difficult
this is to achieve in practice, and similar difficulties are likely to
be experienced in the imaging domain. Some of the difficulties we
forsee are:
\begin{itemize}
\item A real imaging application is likely to utilize a large number
  of ITK filters. It is likely to be a long time before a significant
  portion of ITK is CUDA (or GPU) enabled. Therefore a developer will
  only experience a nett performance gain if the CUDA enabled filter
  offers sufficient speedup to offset memory transfer costs and a
  small proportion of the application is particularly time consuming.
\item Filters that are easy to port to the GPU tend to be fast on CPU
  anyway, and typically don't represent a large proportion of
  application time. Examples include all voxel-wise operations, such
  as masking and arithmetic.
\item Many potentially time consuming operations, such as filtering
  with large kernels, have been highly optimised for CPU
  implementation. It is important the comparisons are made with these
  optimised CPU implementations.
\end{itemize}

We won't discuss the mechanics of CUDA performance profiling in this
article - there are many resources available online.

One point worth noting when testing CUDA enabled ITK filters is that
there is a per-process cost associated with running a CUDA-enabled
application. This cost appears to relate to a number of things,
including loading libraries and initializing the device. This cost can
be very significant - as much as 2.5 seconds on one of our test
machines - and can give an exagerated negative impression of the  filter
performance.

Finally, some positive performance results. These tests were carried
out using a Tesla T10 in a 16 core, 2261.051MHz, Intel L5520  Xeon:
\begin{itemize}
\item Performance improvement of 190 times observed with simple arithmetic, such as adding or subtracting constants from
  images. See {\em simple\_perf\_test.cxx} for details.
\item Image filtering with kernels, for example simple means, offer ??
  for 3d kernels, radius 10 voxels on images size $500 \times 500
  \times 500$. This is potentially interesting for applications
  requiring kernels of specific shape, because most accelerated CPU
  schemes can only implement a restricted range of shape kernels. The
  accelerated CPU scheme used in this test uses a box filtering
  approach that is limited to rectangular filters.
\end{itemize}

\section{Conclusions}
We have provided some simple modifications to ITK infrastructure that
allow integration of CUDA enabled filters with ITK applications, and
provided a number of examples and validation tests. We hope this
framework will encourage immediate experimentation with GPU filters
and inform some of the GPU development scheduled for ITK 4.
